How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
Details in section 2. Started reworking and completing section 3, intermediate check-in.

\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts of at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

@@

\emph{Better performance:} So far it seems like performance of highly dynamic
languages is not actually significantly improved on OO VMs. Jython is around 5
times slower than CPython, for IronPython the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this that the VM's JIT compilers are not prepared to deal with the highly
dynamic behaviour and the complex dispatch semantics of dynamic languages. (XXX
expand)

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe –
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

\item
we use a very expressive \emph{object language} (RPython – an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details – no explicit memory
management, no pieces of C or C-level code.

@@

\item
this meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code – e.g. the garbage collector when
targeting C – is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Our approach – a single ``meta-written'' implementation – naturally leads
to language implementations that have various advantages over the
``hand-written'' implementations.  First of all, it is a single-source
approach – we explicitly seek to solve the problem of proliferation of
implementations.  In the sequel we will show more precise evidence that
this can be done in a practical way with no major drawback.  By itself
this would already be a valid argument against the need for
standardization on a single OO VM.  But there are also other advantages
in generating language implementations. In our opinion these are very
significant, to the extent that it hints that meta-programming, though
not widely used in general-purpose programming, is an essential tool in
a language implementer's toolbox.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
– a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking \cite{D12.1}, and general
performance improvements found by extensive experimentation \cite{D06.1}, some
of which were back-ported to CPython.

If we compare with hand-writing an implementation for a specific OO VM,
then the latter requires not only good knowledge of the OO VM in
question and its object model – it requires the language implementer to
fit the language into the imposed models.  For example, it is natural to
map classes and instances of a dynamic object-oriented language to the
OO VM's notion of classes and instances, but this might not be a simple
task at all if the models are substantially different and/or if the OO
VM is essentially less dynamic than the language to implement.  In our
approach, this efforts is done at two levels: in a first step, while
writing the interpreter, the implementer does not need to worry about
integration with an OO VM's object model.  Of course, the integration
effort does not simply vanish – indeed, a simple translation of such an
interpreter to a given OO VM would give an interpreter for the dynamic
language which is unable to communicate with the host VM (which might
already be interesting in specific cases, but not in general).
Integration comes as a second step, and occurs at a different level, by
introducing mappings between the relevant classes of the interpreter and
the corresponding classes of the OO VM.  As of yet we have no evidence
that this makes the total integration effort much lower or higher; the
point is that it has proven possible to take the single, OO
VM-independent source code of PyPy's Python interpreter, and produce
from it a version that runs in and integrates with the rest of the .NET
environment – by writing \emph{orthogonal} code.

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By ``aspect'' we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically – in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional ``stackless'' transformation is
able to turn the Python interpreter – also written in a simple highly
recursive style – into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.

>>>>

\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

@@

\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. Jython
\footnote{Python on the Java VM} is around 5
times slower than CPython, for IronPython\footnote{Python on .NET, which
gives up on some features to improve performance}
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are not prepared to deal with the highly
dynamic behaviour and the complex dispatch semantics of dynamic languages. (XXX
expand)

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

\item
we use a very expressive \emph{object language} (RPython -- an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details -- no explicit memory
management, no pieces of C or C-level code.

@@

\item
this meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code -- e.g. the garbage collector when
targeting C -- is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Our approach -- a single ``meta-written'' implementation -- naturally leads
to language implementations that have various advantages over the
``hand-written'' implementations.  First of all, it is a single-source
approach -- we explicitly seek to solve the problem of proliferation of
implementations.

Separating the implementation of a language in a high-level
``description'' and a custom translation framework has also many
advantages -- in our opinion significant enough to hint that
meta-programming, though not widely used in general-purpose programming,
is an essential tool in a language implementer's toolbox.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
-- a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking \cite{D12.1}, and general
performance improvements found by extensive experimentation \cite{D06.1}, some
of which were back-ported to CPython.

\subsection{The effort of writing a translation toolchain}

Of course, the price to pay is the need for a translation toolchain
capable of analyzing and transforming the high-level source code and
generating lower-level output in various languages.  Of course, the
translation toolchain, once written, can be reused to implement other
languages as well.  Even so, we found that the required effort that must
be put into the translation toolchain in the first place is still much
lower than that of writing a complete, commercial-quality OO VM.  A
reason appears to be that we could design our translation toolchain
specifically for our needs, i.e. a language implementer's needs, instead
of for general-purpose usage.

...

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By ``aspect'' we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically -- in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional ``stackless'' transformation is
able to turn the Python interpreter -- also written in a simple highly
recursive style -- into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.

\subsection{Integration with a host OO VM}

A good example of this is to compare it against the task of hand-writing
an implementation for a specific, choosen OO VM.  The latter requires
not only good knowledge of the OO VM in question and its object model --
it requires the language implementer to fit the language into the
imposed models.  For example, it is natural to map classes and instances
of a dynamic object-oriented language to the OO VM's notion of classes
and instances, but this might not be a simple task at all if the models
are substantially different and/or if the OO VM is essentially less
dynamic than the language to implement.  In our approach, this task
is done at two levels: in a first step, while writing the interpreter,
the implementer does not need to worry about integration with an OO VM's
object model.  Of course, the integration effort does not simply vanish
-- indeed, a simple translation of such an interpreter to a given OO VM
would give an interpreter for the dynamic language which is unable to
communicate with the host VM (which might already be interesting in
specific cases, but not in general).  Integration comes as a second
step, and occurs at a different level, by introducing mappings between
the relevant classes of the interpreter and the corresponding classes of
the OO VM.  As of yet we have no evidence that this makes the total
integration effort much lower or higher; the point is that it has proven
possible to take the single, OO VM-independent source code of PyPy's
Python interpreter, and produce from it a version that runs in and
integrates with the rest of the .NET environment -- by writing
\emph{orthogonal} code.
